---
title: "BIOF Final Project"
author: "Nicholas Giordano"
date: "5/21/2019"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(foreign)
library(ggplot2)
library(dplyr)
library(lmerTest)
library(lme4)
library(tidyr)
library(dplyr)
library(tableone)
library(sjPlot)
library(plotrix)
```

## Objective

Identify if there is an association between early pain management in the battlefield and long-term average pain scores after combat injury

```{r, echo=FALSE}
df <- read.csv("ngdata.csv")
```

## Describe the sample

The dataset includes:


 1. **ID** = study number
 2. **Pain** = Pain scores, 0 to 10 higher scores indiciate greater pain 
 3. **Intervention** = Early pain management via regional anesthesia (RA) or no early pain management (No Early RA)
 4. **Month** = Month since injury that pain score was documented
 5. **LOS** = Length of stay in days at inital hospital as an indiciator of severity of injury
 6. **Injury** = Injury Severiy Score, 0 to 75 higher scores indiciate greater injury severity 
 7. **Observations** = # of observations in the study per subject
 

\pagebreak

```{r, echo=FALSE}
df2 <- df %>% 
  group_by(ID) %>% 
  mutate(observations = n()) %>% 
  ungroup() #AD Should always ungroup if you're using the dataset later
summary(df2)
```

### Sample Described by Intervention Group 

*Individuals with early pain management have slighly lower pain scores*

```{r, echo=FALSE}
table1 <- CreateTableOne(
  vars=c("Pain", "Month", "LOS", "Injury", "observations"),
  data=df2,
  strata="Intervention"
)

#AD Specify the statistical test that is being run here. I'm only seeing p-values

print(table1)
```



```{r, include=FALSE}
df3 <- df2 %>% 
  group_by(Intervention, Month) %>% 
  summarize(avPain = mean(Pain))
  
 df4 <- df2 %>%
  group_by(Intervention, Month) %>% 
  summarize(sdPain=std.error(Pain))
 
df5 <- merge(df3,df4, by=c("Intervention", "Month"))

#AD This process can be done in one statement:
#AD
#AD df5 <- df2 %>% 
#AD    group_by(Intervetion, Month) %>% 
#AD    summarize(avPain = mean(Pain), sdPain = std.error(Pain))

```

## Plot Average Pain Scores over Time 

There is no clear difference in pain scores by individuals who did and did not receive early pain management over time. Need to adjust for other factors to see if the difference observed in the table above is meaningful or not.

```{r, echo=FALSE}
ggplot(df5, aes(x=Month, y=avPain, linetype=Intervention)) + 
  geom_point() + 
  geom_line() + 
  geom_errorbar(aes(ymin=avPain-sdPain, ymax=avPain+sdPain), width=.2) +
  ylim(0,5) + 
  # scale_x_continuous(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) + #AD There is a shortcut
  scale_x_continuous(breaks = 1:24) + 
  ggtitle("Average Pain Per Month By Intervention") + 
  theme_classic()
```
```{r, include=FALSE}
model1 <- lmer(Pain ~ Intervention +Month + LOS + Injury + (ID | Month), data=df2)
summary(model1)

#AD Might want to re-think this. The usual random effect here would be (1|Month), 
#AD i.e., a random intercept model. It's also not clear to me why you need a mixed model or a 
#AD hierarchical model in this case. The individuals were selected independent of Month, not nested within Month, since Month is time since injury. 
```

##Adjusted model
When we adjust for clinical factors, such as length of stay and injury severity, on average early pain management is associated with lower pain scores over time
```{r,echo=FALSE}

plotx<-plot_model(model1, type = "pred", terms = c("Month [1,2,3,4,5,6,9,12,15,18,21,24]", "Intervention [No Early RA,Early RA]"))+
    ggtitle("Estimated Marginal Means From 1-24 Months") + labs( y="Average Pain (0-10)", x="Time Since Injury to Observation (1-24)")+
      theme_bw()+
  theme(axis.line = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    legend.position  = c(0.75,0.75),
    panel.background = element_blank()) + 
  ylim(0,5) +
  scale_x_continuous(limits = c(1, 24), breaks= c(1,6,12,18,24))
plotx
```

```{r, eval = F}

#AD This is for my understanding of what plot_model does here. Maybe helps your 
#AD understanding too. Don't trust a black box until you can verify at least some of it. 

fixed_effects <- fixef(model1)
newdata <-  as.data.frame(model.matrix(model1)) %>% mutate(LOS = mean(LOS), Injury = mean(Injury)) %>% as.matrix()
predictions <- newdata %*% fixed_effects
augmented_data <- df2 %>% mutate(pred = predictions)

plot_data <- augmented_data %>% group_by(Intervention, Month) %>% summarize(m = mean(pred))

ggplot(plot_data, aes(x = Month, y = m, group = Intervention, color = Intervention)) + 
  geom_line() + coord_cartesian(y = c(0, 5)) +
  theme_classic() +
  theme(legend.position = c(.8, .8)) + 
  labs(title = 'Estimated Marginal Means from 1-24 Months',
       x = 'Time Since Injury to Observation (1-24)',
       y = 'Average Pain (1-10)')
  
```

