---
title: "BIOF Final Project"
author: "Nicholas Giordano"
date: "5/21/2019"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(foreign)
library(ggplot2)
library(dplyr)
library(lmerTest)
library(lme4)
library(tidyr)
library(dplyr)
library(tableone)
library(sjPlot)
library(plotrix)
```

## Objective

Identify if there is an association between early pain management in the battlefield and long-term average pain scores after combat injury

```{r, echo=FALSE}
df <- read.csv("ngdata.csv")
```

##Describe the sample
The dataset includes:


 1. **ID** = study number
 2. **Pain** = Pain scores, 0 to 10 higher scores indiciate greater pain 
 3. **Intervention** = Early pain management via regional anesthesia (RA) or no early pain management (No Early RA)
 4. **Month** = Month since injury that pain score was documented
 5. **LOS** = Length of stay in days at inital hospital as an indiciator of severity of injury
 6. **Injury** = Injury Severiy Score, 0 to 75 higher scores indiciate greater injury severity 
 7. **Observations** = # of observations in the study per subject
 

\pagebreak

```{r, echo=FALSE}
df2<-df %>% 
  group_by (ID) %>% 
  mutate(observations =n())
summary(df2)
```

###Sample Described by Intervention Group 
*Individuals with early pain management have slighly lower pain scores*
```{r, echo=FALSE}
table1<-CreateTableOne(
  vars=c("Pain", "Month", "LOS", "Injury", "observations"),
  data=df2,
  strata="Intervention"
)
print(table1)
```

```{r, include=FALSE}
df3<-df2 %>% 
  group_by(Intervention, Month) %>% 
  summarize(avPain=mean(Pain))
  
 df4<-df2 %>%
  group_by(Intervention, Month) %>% 
  summarize(sdPain=std.error(Pain))
 
df5<-merge(df3,df4, by=c("Intervention", "Month"))

```

##Plot Average Pain Scores over Time 
There is no clear difference in pain scores by individuals who did and did not receive early pain management over time. Need to adjust for other factors to see if the difference observed in the table above is meaningful or not.

```{r, echo=FALSE}
ggplot(df5, aes(x=Month, y=avPain, linetype=Intervention)) + 
  geom_point() + 
  geom_line() + 
  geom_errorbar(aes(ymin=avPain-sdPain, ymax=avPain+sdPain), width=.2) +
  ylim(0,5) + 
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24)) +
  ggtitle("Average Pain Per Month By Intervention") + 
  theme_classic()
```
```{r, include=FALSE}
model1 <- lmer(Pain ~ Intervention +Month + LOS + Injury + (ID | Month), data=df2)
summary(model1)
```

##Adjusted model
When we adjust for clinical factors, such as length of stay and injury severity, on average early pain management is associated with lower pain scores over time
```{r,echo=FALSE}

plotx<-plot_model(model1, type = "pred", terms = c("Month [1,2,3,4,5,6,9,12,15,18,21,24]", "Intervention [No Early RA,Early RA]"))+
    ggtitle("Estimated Marginal Means From 1-24 Months") + labs( y="Average Pain (0-10)", x="Time Since Injury to Observation (1-24)")+
      theme_bw()+
  theme(axis.line = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    legend.position  = c(0.75,0.75),
    panel.background = element_blank()) + 
  ylim(0,5) +
  scale_x_continuous(limits = c(1, 24), breaks= c(1,6,12,18,24))
plotx
```