---
title: "Differential expression analysis of cone and S-cone-like photoreceptors"
Date: 05-22-2019
output:
  html_document:
    df_print: paged
Author: Samantha Papal
---

## Background
- Cone photoreceptors are necessary for visual acuity and color perception in human.
- Cones rely on rod photorecptor for their survival.
- Loss of rods leads to subsequent death of cones which can lead to blindness
- We compared purified cone photoreceptors from WT mice with S-cone-like cells from a    knockout mouse line (*Nrl*^-/-^) where cone photoreceptors survive in the absence of      rods.
- Hypothesis: Some genes expressed by *Nrl*^-/-^ mice are responsible for their      maintenance in the absence of rods



```{r, include= FALSE}
setwd("~/Documents/FAES/Visualization_with_R/Samantha_Papal_Data_visualization_with_R/Data_visualization")
getwd()
```
##Principal component analysis plot


1. Load libraries
```{r, message= FALSE, warning= FALSE}
library(tidyverse)
library(DESeq2)
library(ggplot2)
```
2. Load DESeq2 data
```{r}
Normalized_data_for_PCA <- readRDS("Normalized_data_for_PCA")
```
3. Create the PCA plot
```{r}
# Create labels to be used in legend
labelsName <- c("cone P2", "cone P4", "cone P6", "cone P10", "S-cone-like P2", "S-cone-like P4", "S-cone-like P6", "S-cone-like P10")
```
```{r}
# Gete the PCA data to plot with ggplot
pcadata <- plotPCA(Normalized_data_for_PCA, returnData = TRUE)
percentVar <- round(100*attr(pcadata,"percentVar"))
```
```{r}
# Plot PCA
ggplot(pcadata, aes(PC1,PC2, color = fct_relevel(condition, c("control_P2", "control_P4", "control_P6", "control_P10", "experiment_P2", "experiment_P4", "experiment_P6", "experiment_P10")))) +
  geom_point(size = 3) +
  labs(x = paste0("PC1: ",percentVar[1],"% variance"), 
       y = paste0("PC2: ",percentVar[2],"% variance"), 
       title = "Principal component analysis of cone and S-cone-like photoreceptors", 
       caption = "Cone data:Daum et al., 2017\nS-cone-like data: Kim et al., 2016") +
  coord_fixed() + 
  scale_color_discrete(name = "Samples", labels = labelsName) +
  theme(panel.background = element_blank(), 
        axis.line = element_line(), 
        panel.grid.major = element_line(colour = alpha("black", 0.2)), 
        legend.key = element_rect(colour = "transparent", fill = "transparent"), 
        plot.caption = element_text(size = 6),
        plot.title = element_text(size = 14, margin = margin(0,0,40,0), hjust = 0.3))
```

##Hierachical clustering of cone specific genes in Cone and S-cone-like cells

1. Load library
```{r, message= FALSE, warning= FALSE}
library(gplots)
```
2. Load file containing the selected genes to plot
```{r, message= FALSE}
# Load file
genes_meta <- read_csv("Table_cone_genes.csv")
```
```{r, warning= FALSE}
# Organize the table for the heatmap and dconvert it to a matrix for clustering analysis
heatmap_meta <- genes_meta[, c(2:10)]
rownames(heatmap_meta) <- heatmap_meta$mgi_symbol
heatmap_meta <- select(heatmap_meta,-mgi_symbol)
heatmap_meta <- as.matrix(heatmap_meta)
```
```{r}
# Names of the colums for the heatmap
colnames(heatmap_meta) <- c("cone P2", "cone P4", "cone P6", "cone P10", "cone-like P2", "cone-like P4", "cone-like P6", "cone-like P10") # rename samples
```
```{r warning= FALSE}
# Clustering options
row_distance = dist(heatmap_meta, method = "euclidean")
row_cluster = hclust(row_distance, method = "ward.D2")

# Sequential color from http://colorbrewer2.org/
colors <- colorRampPalette(c('#ffffcc','#ffeda0','#fed976','#feb24c','#fd8d3c','#fc4e2a','#e31a1c','#bd0026','#800026'))

# Generate heatmap
par(cex.main=1)
heatmap.2(round(heatmap_meta, digits = 2),
          
          # dendrogram control
          Rowv = as.dendrogram(row_cluster), # apply clustering method
          dendrogram = "row", # generate only row dendrogram
          Colv = NULL, # No column dendrogram
          
          # colors
          col = colors, # use colors defined previously
          
          # level trace
          trace = "none", # turn off trace lines inside heat map
          
          # 2gins and row/column label sizes
          margins = c(6,5), # widens margins around plot
          cexRow = 1, # font size for row labels dependent on number of rows
          cexCol = 1, # font size for column labels dependent on number of columns
          cex.main = 2,
          
          # color key and density information
          key = TRUE, # plot color key
          keysize = 0.4, # color key size
          key.par = list(cex=0.6),
          density.info = "none", # turn off density plot inside color legend
          key.title = "NULL",
          key.xlab = expression(log[2]~(CPM+1)),
          key.ylab = "",
          
          # plot labels
          main = "Expression pattern of cone specific genes",
          xlab = "", # x-axis title
          ylab = NULL, # y-axis title
          srtCol = 90,
          adjCol = c(1, 0.5),
          labCol = c("cone\nP2", "cone\nP4", "cone\nP6", "cone\nP10", "S-cone-like\nP2", "S-cone-like\nP4", "S-cone-like\nP6", "S-cone-like\nP10"),
          
          lhei = c(4, 14),
          lwid = c(3, 14))

```

## Gene enrichment analysis

1. Load library
```{r, message= FALSE, warning= FALSE}
library(gProfileR)
```
2. Open tables to plot the graph
```{r, message= FALSE, warning= FALSE}
DE_gprofileR_results <- read_csv("DE_gprofileR_results.csv")
domain.code <- read_csv("domain.code.csv")
```
3. Plot Human Phenotype Ontology for the differentialy expressed genes
```{r fig.asp= 0.8}
  Data <- filter(DE_gprofileR_results, domain == "hp")
 ggplot(Data, aes(x = query.number, y = term.name, size = overlap.size, color = p.value)) + 
   geom_point() + 
   scale_x_discrete(labels = element_blank()) +
    labs(title = "Human Phenotype Ontology of S-cone like enriched genes",
         color = "p.value", size = "count") +
   guides(color = guide_colorbar(order = 1), 
          size = guide_legend(order = 0)) +
    theme(panel.background = element_blank(), 
          axis.line.y = element_line(), 
          axis.title.x = element_blank(), 
          axis.title.y = element_blank(), 
          axis.text.y = element_text(size=11), 
          plot.title = element_text(size = 14, margin = margin(0,0,30,0), hjust = 0.7))
  
```