---
title: "A quick tour of R"
author: Abhijit Dasgupta, PhD
date: Spring 2019

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F, comment = '#> ', cache=FALSE)
```

layout: true

<div class="my-header">
<span>BIOF 439, Spring 2019</span></div>

---

## The bare essentials

- Everything in R is an object with a name
- Things are nouns
    - Nouns in R are _data objects_, like scalars, matrices, data.frames/tibbles, strings, vectors
- Nouns are acted upon by verbs
    - Verbs in R are _functions_, like `mean(x)`, `nrow(d)`, `dim(d)`, `ggplot` and so on
- You can modify verbs with adverbs
    - Adverbs in R are _function options_, like `mean(x, na.rm=T)`, `geom_point(color='green')`

---

## The bare essentials
- You have to name things to store them
    - This is done with the `<-` operator, e.g. 
        - `mn <- mean(x, na.rm=T)` stores the result of the average
        - `my_theme <- function() theme_bw() + theme(axis.title=element_text(size=14))` stores the function, which you'll call as `my_theme()`

> You can see the objects you have created either by typing `ls()` in the console, or looking in the Environment pane

> Note, built-in objects don't show up in the Environment pane or using `ls()`

---

## The bare essentials

There are three kinds of brackets in R

`[]` are used for extracting elements from arrays, matrices, data frames.

- `x[3]` is the 3rd element of an array `x`
- `d[1,3]` is the element in the 1st row and 3rd column of a matrix/data frame `d`
- `d[2,]` is the entire first __row__ of a matrix/data frame `d`
- `d[,4]` is the entire 4th __column__ of a matrix/data frame `d`
    
---

## The bare essentials

There are three kinds of brackets in R

`()` are used for specifying arguments to functions

- `mean(x)` gives the mean of an array of numbers `x`
- `summary(d)` gives a summary representation of a data frame `d`

---

## The bare essentials

There are three kinds of brackets in R

`{}` are used to contain groups of commands/statements

.pull-left[

A conditional statement

```{r Rintro-1, eval = F}
if (age < 18){
   person <- 'Minor'
} else if (age > 65) {
  person <- 'Senior'
} else {
  person <- 'Adult'
}
```
]
.pull-right[

A function definition

```{r Rintro-2, eval=F}
my_mean <- function(x, na.rm = T){
  if(na.rm){
    x <- x[!is.na(x)]
  }
  s <- sum(x)
  n <- length(x)
  mn <- s / n # There is a built-in function mean, so I don't use that word
  return(mn)
}
```

]

---

## Data types

A __scalar:__

- `29`
- `"cherry"`
- `TRUE`

---

## Data types

### A __scalar:__

- `29` : _numeric_
- `"cherry"`: _character_
- `TRUE`: _logical_

---

## Data types

### Vectors/Arrays

These are constructed using the `c()` function (for _concatenate_).

```{r Rintro-3}
c(1,2,5,6,7,8)
c('apple','berry','melon','citrus')
```

> Vectors must all contain objects of the same type. Can't mix and match


---

## Data types

### Matrices (2-d arrays)

These are typically built from vectors

```{r Rintro-4}
x <- c(1,2,4,5,6,7)
y <- 10:16  # Shortcut for c(10,11,12,13,14,15,16)
```

.pull-left[
```{r Rintro-5, eval=T}
cbind(x, y) # Vectors as columns
```
]
.pull-right[
```{r Rintro-6, eval=T}
rbind(x, y) # Vectors as rows
```
]

---

## Data types

### Lists

Lists are basically buckets or containers. Each element of a list can be anything, even other lists

.pull-left[
```{r b1, eval=F}
my_list <- list('a', c(2,3,5,6), head(ggplot2::mpg))
my_list
```
]
.pull-right[
```{r Rintro-7, echo=F, eval=T, ref.label='b1'}

```
]
---


## Data types

### data.frame/tibble

This is the typical container for tabular data

- must be rectangular
- each column can be of a different type
- elements within each column have to be of the same type

---

## Data types

### data.frame/tibble

```{r Rintro-8}
beaches <- rio::import('data/sydneybeaches3.csv') # use the import function from the package rio
class(beaches)
dim(beaches)
head(beaches)

```


---

## Data types

### data.frame/tibble

```{r Rintro-9, message=F}
library(tidyverse) # Activate the tidyverse package
beaches_t <- as_tibble(beaches)
class(beaches_t)
beaches_t
```

---

## Data types

### data.frame/tibble

Extracting columns from a data frame:

1. `beaches$temperature`
1. `beaches[,'temperature']`
1. `beaches[['temperature']]`
1. `beaches[,7]`
1. `beaches[[7]]`
---

## Packages in R

R is a modular environment with some base functionality that is augmented by __packages__ (think of them as modules)

- Packages can contain _functions_ and _data_
- There are over 15K packages on CRAN, the Comprehensive R Archive Network
- There are over 1600 packages on Bioconductor, the main repository for bioinformatics resources
    - Analytic, Annotation, Experimental data and Workflow packages

--

#### Finding packages

1. CRAN [Task views](https://cran.r-project.org/web/views/){target=_blank}
1. Bioconductor [BiocViews](http://bioconductor.org/packages/release/BiocViews.html){target=_blank}
1. GitHub (open source collaboration and version control environment)

---

## Installing packages

.pull-left[
### From CRAN
```{r Rintro-10, eval=F}
install.packages("tidyverse")
```

### From Bioconductor
```{r Rintro-11, eval = F}
install.packages("BiocManager") # do once 
BioManager::install('limma')
```

### From GitHub

```{r Rintro-12, eval = F}
install.packages('remotes') # do once
remotes::install_github("rstudio/rmarkdown") # usual format is username/packagename
```
]
.pull-right[
> GitHub often hosts development version of packages published on CRAN or Bioconductor

> Both CRAN and Bioconductor have stringent checks to make sure packages can run properly, with no obvious program flaws. There are typically no
  guarantees about analytic or theoretical correctness, but most packages have been crowd-validated and there are several reliable developer groups
  including RStudio
]

---

## Using packages

You have to first "activate" the package in your current working session using the `library` function.

```{r Rintro-13, include=F}
detach_pkg <- search()[str_detect(search(), 'gg')]
map(detach_pkg, function(x) {
  ind <- which(str_detect(search(), x))
  if(length(ind) > 0) detach(pos = ind)
})

```

.pull-left[
```{r Rintro-14, error=TRUE}
ggplot(beaches, aes(temperature, rainfall)) +
  geom_point()
```
]

--

.pull-right[
```{r Rintro-15, message=F, warning = F, fig.height=4}
library(ggplot2) # or library(tidyverse)
ggplot(beaches, aes(temperature, rainfall)) +
  geom_point()
```

]

---
class: center, middle

# Tidying data using the tidyverse

---

## What is the ["Tidyverse"](http://www.tidyverse.org)?



<div style="display:flex;align-items:center;font-size:30pt;font-family:serif;width:100%;height:300px;background-color:wheat;text-align:center; border: 1px solid red; position: relative;">

The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures

</div>

---

## What is the ["Tidyverse"](http://www.tidyverse.org)?

A set of R packages that:

<ul>
  <li> help make data more computer-friendly
  <li>while making your code more human-friendly
</ul>

- Most of these packages are (co-)written by Dr. Hadley Wickham, who has rockstar status in the R world
- They are supported by the company RStudio

---
class:middle, center

# Tidying data

---

## Tidy data

<div style="display:flex;align-items:center;font-size:30pt;font-family:sans-serif;width:100%;height:300px;background-color:wheat;text-align:center; padding-left:30px;border: 1px solid red; position: relative;">

Tidy datasets are all alike, <br/>
but every messy data is messy in its own way

</div>

---

## Tidy data

Tidy data is a **computer-friendly** format based on the following characteristics:

- Each row is one observation
- Each column is one variable
- Each set of observational unit forms a table

All other forms of data can be considered **messy data**.

---

## Let us count the ways

There are many ways data can be messy. An incomplete list....

+ Column headers are values, not variables
+ Multiple variables are stored in a single column
+ Variables are stored in both rows and columns
+ Multiple types of observational units are saved in the same table
+ A single observational unit is stored in multiple tables

---

## Ways to have messy (i.e. not tidy) data

1. Column headers contain values

Country   |   < $10K    | $10-20K    | $20-50K   | $50-100K    | > $100K
----------|-------------|------------|-----------|-------------|---------
India     |   40        |  25        |   25      |  9          |  1
USA       |   20        |  20        |  20       | 30          |  10

---

## Ways to have messy (i.e. not tidy) data

Column headers contain values

Country   |   Income  | Percentage
----------|-----------|------------
India     |  < $10K   |  40
USA       |  < $10K   | 20

This is a case of reshaping or melting 

---

## Ways to have messy (i.e. not tidy) data

Multiple variables in one column

Country  | Year   | M_0-14  | F_0-14  | M_ 15-60  | F_15-60  | M_60+  | F_60+
---------|--------|---------|---------|-----------|----------|--------|-------
UK       |  2010  |         |         |           |          |        | 
UK       |  2011  |         |         |           |          |        | 

<p>
Separating columns into different variables

Country  | Year   | Gender  | Age    | Count
---------|--------|---------|--------|-------




---

## Tidying data

The typical steps are 

+ Transforming data from wide to tall (*gather*) and from tall to wide (*spread*)
+ Separating columns into different columns
+ Putting columns together into new variables

---
class: middle, center

# Cleaning data

---

## Some actions on data

+ Creating new variables (*mutate*)
+ Choose some columns (*select*)
+ Selecting rows based on some criteria (*filter*)
+ Sort data based on some variables (*arrange*)

---

## Example data

```{r Rintro-16, echo = T}
head(mtcars, 3)
```


- Car names are in an attribute of the data.frame called `rownames`. So it's not in a column
- We might want to convert fuel economy to metric 
- We might just want to look at the relationship between displacement and fuel economy based on number of cylinders

---

## Example data ([link](https://dl.dropboxusercontent.com/s/pqavhcckshqxtjm/brca.csv))

```{r Rintro-17, error=T}
link <- 'https://dl.dropboxusercontent.com/s/pqavhcckshqxtjm/brca.csv'
brca_data <- rio::import(link)
```
<img src = "https://dl.dropboxusercontent.com/s/k3r12s7d22zhku8/Screenshot%202018-09-26%2000.27.50.png" 
  width="600" height="150"></img>

---

## The `tidyverse` package

The `tidyverse` package is a meta-package that installs a set of packages that are useful for data 
cleaning, data tidying and data munging (manipulating data to get a computationally "attractive" 
dataset)

---

## The `tidyverse` package

```{r Rintro-18, warning=FALSE}
# install.packages('tidyverse')
library(tidyverse)
```


> You can specify a function from a particular package as `dplyr::filter`. Note there
are two colons there


---

## Core `tidyverse` packages

```{r Rintro-19, echo = FALSE, message=F, warning=F}
my_tbl <- tibble::tribble(
  ~Package, ~Description,
      'ggplot2',     'Data visualization',
      'tibble',     'data.frame on steroids',
      'tidyr',     'Data tidying (today)',
      'readr',     'Reading text files (CSV)',
      'purrr',     'Applying functions to data iteratively',
      'dplyr',     'Data cleaning and munging (today)',
      'stringr',     'String (character) manipulation',
      'forcats',     'Manipulating categorical variables'
  )

require(knitr)
kable(my_tbl, digits = 3, row.names = FALSE, align = "l",
              caption = NULL, format='html')
```

---

## Additional `tidyverse` packages

```{r Rintro-20, echo=F}
my_tbl <- tibble::tribble(
  ~Package, ~Description,
      'readxl',     'Read Excel files',
      'haven',     'Read SAS, SPSS, Stata files',
      'lubridate',  'Deal with dates and times'  ,
      'magrittr',     'Provides the pipe operator %>%',
      'glue', 'Makes pasting text and data easier')

require(knitr)
kable(my_tbl, digits = 3, row.names = FALSE, align = "l",
              caption = NULL, format='html')

```

### Additional useful packages

```{r Rintro-21, echo=F}
my_tbl <- tibble::tribble(
  ~Package, ~Description,
  'broom', 'Turns the results of models or analysis into tidy datasets',
  'fs','Allows directory and file manipulation in OS-agnostic manner',
  'here', 'Allows robust specification of directory structure in a Project')

require(knitr)
kable(my_tbl, digits = 3, row.names = FALSE, align = "l",
              caption = NULL, format='html')
```

---
class: middle, center

# Pipes

---

## Pipes

Pipes (denoted `%>%`, spoken as "then") are to analytic pipelines as `+` is to `ggplot` layers

```{r a2, eval=T}
mpg1 <- mpg %>% mutate(id=1:n()) %>% select(id, year, trans, cty, hwy)
mpg_metric <- mpg1 %>% 
  mutate_at(vars(cty, hwy), function(x) {x * 1.6/3.8})
```

.pull-left[

Original data

```{r Rintro-22, echo=F}
knitr::kable(head(mpg1, 5), format='html')
```

]
.pull-right[

Transformed data
```{r Rintro-23, echo = F}
knitr::kable(head(mpg_metric, 5), format='html')
```

]

--

> Note I'm assigning a name to the transformed data. Otherwise it'll be lost

---

## Verbs to use in pipes

The verbs in `tidyverse` are specially useful in pipes

```{r Rintro-24, echo=F}
my_tbl <- tribble(
  ~"Verb", ~"Functionality",
  'mutate','Transform a column with some function',
  'select', 'Select some columns in the data',
  'arrange', 'Order the data frame by values of a column(s)',
  'filter','Keep only rows that meet some data criterion',
  'group_by', 'Group by levels of a variable',
  'gather', 'Transform a wide dataset to a long dataset',
  'spread','Transform a long dataset to a wide dataset',
  'separate', 'Separate one column into several columns',
  'unite', 'Concatenate several columns into 1 column')
knitr::kable(my_tbl, align = 'l', format='html')
```

Pipes almost always start with a data.frame/tibble object, and then "pipes" that data through different transformations (functions)

At each `%>%`, the results of the previous step are used as input for the next step.

---
class: middle, center

# A complicated example

---

## Grab the raw data

```{r ex1, eval=F}
url <- "http://varianceexplained.org/files/Brauer2008_DataSet1.tds"
raw_data <- read_delim(url, delim='\t')
head(raw_data)
```
```{r Rintro-25, echo=F, eval=T, ref.label='ex1'}

```


---

## Look at the annotation data

```{r ex2, eval=F}
head(raw_data$NAME)
```
```{r Rintro-26, ref.label='ex2',echo=F, eval=T}

```


---

## Separate annotation into columns

```{r ex3, message=FALSE, eval=F}
cleaned_data <- raw_data %>% 
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), #<<
           sep = "\\|\\|")#<<
head(cleaned_data)
```
```{r Rintro-27, eval=T, echo=F, ref.label='ex3'}

```


---

## Get rid of padding in annotation

```{r ex4, eval=F, message=FALSE}
cleaned_data <- raw_data %>% 
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), 
           sep = "\\|\\|") %>% 
  mutate_at(vars(name:systematic_name), funs(stringr::str_trim)) #<<
head(cleaned_data)
```
```{r Rintro-28, echo=F, eval=T, ref.label='ex4'}

```


---

## Get rid of some columns

```{r ex5, eval=F, message=FALSE}
cleaned_data <- raw_data %>% 
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), 
           sep = "\\|\\|") %>% 
  mutate_at(vars(name:systematic_name), funs(stringr::str_trim)) %>% 
  select(-number, -GID, -YORF, -GWEIGHT) #<<
head(cleaned_data)
```
```{r Rintro-29, ref.label='ex5', eval=T, echo=F, message=F}

```


---

## Make data tidy

```{r ex6, eval=F,  message=FALSE}
cleaned_data <- raw_data %>% 
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), 
           sep = "\\|\\|") %>% 
  mutate_at(vars(name:systematic_name), funs(stringr::str_trim)) %>% 
  select(-number, -GID, -YORF, -GWEIGHT) %>% 
  tidyr::gather(sample, expression, G0.05:U0.3) #<<
head(cleaned_data)
```
```{r Rintro-30, ref.label='ex6', echo=F, eval=T, message=F}

```


---

## Split columns

```{r ex7, eval=F, message=FALSE}
cleaned_data <- raw_data %>% 
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), 
           sep = "\\|\\|") %>% 
  mutate_at(vars(name:systematic_name), funs(stringr::str_trim)) %>% 
  select(-number, -GID, -YORF, -GWEIGHT) %>% 
  tidyr::gather(sample, expression, G0.05:U0.3) %>% 
  separate(sample, c("nutrient", "rate"), sep=1, convert = TRUE) #<<
head(cleaned_data)
```
```{r Rintro-31, ref.label='ex7', echo=F,eval=T, message=F}

```


---

## Get rid of rows with missing expression or name

```{r ex8, eval=F, message=FALSE}
cleaned_data <- raw_data %>% 
  separate(NAME, c("name", "BP", "MF", "systematic_name", "number"), 
           sep = "\\|\\|") %>% 
  mutate_at(vars(name:systematic_name), funs(stringr::str_trim)) %>% 
  select(-number, -GID, -YORF, -GWEIGHT) %>% 
  tidyr::gather(sample, expression, G0.05:U0.3) %>% 
  separate(sample, c("nutrient", "rate"), sep=1, convert = TRUE) %>% 
  filter(!is.na(expression), systematic_name != '') #<<
head(cleaned_data)
```
```{r Rintro-32, ref.label='ex8', eval=T, echo=F, message=F}

```


---

## Visualize


```{r ex9, eval=T, fig.height=3, fig.width=6, fig.pos='center'}
cleaned_data %>%
  filter(BP == "leucine biosynthesis") %>%
  ggplot(aes(rate, expression, color = nutrient)) + #<<
  geom_point() +#<<
  geom_smooth(method = "lm", se = FALSE) + #<<
  facet_wrap(~name + systematic_name, nrow=1) + #<<
  theme(legend.position='right') #<<

```

---
class: middle, center

# Modeling and the broom package

---

# Representing model relationships

In R, there is a particularly convenient way to express models, where you have

- one dependent variable
- one or more independent variables, with possible transformations and interactions

```
y ~ x1 + x2 + x1:x2 + I(x3^2) + x4*x5
```

--

`y` depends on ...
--

- `x1` and `x2` linearly
--

- the interaction of `x1` and `x2` (represented as `x1:x2`)
--

- the square of `x3` (the `I()` notation ensures that the `^` symbol is interpreted correctly)
--

- `x4`, `x5` and their interaction (same as `x4 + x5 + x4:x5`)

---

# Representing model relationships

```
y ~ x1 + x2 + x1:x2 + I(x3^2) + x4*x5
```

This interpretation holds for the vast majority of statistical models in R

- For decision trees and random forests and neural networks, don't add interactions or transformations, since the model will try to figure those out on their own

---

# Our first model

```{r Rintro-34, echo=T}
library(survival)
data(pbc)
myLinearModel <- lm(chol ~ bili, data = pbc)
```

Note that everything in R is an **object**, so you can store a model in a variable name.

This statement runs the model and stored the fitted model in `myLinearModel`

R does not interpret the model, evaluate the adequacy or appropriateness of the model, or comment on whether looking at the relationship between cholesterol and bilirubin makes any kind of sense.

---

# Our first model

```{r Rintro-35, echo=T}
myLinearModel
```

> Not very informative, is it?

---

# Our first model

```{r Rintro-36, echo=T}
summary(myLinearModel)
```

> A little better

???

Talk about the different metrics in this slide


---

# Our first model

```{r Rintro-37, echo=T}
broom::tidy(myLinearModel)
```

```{r Rintro-38, echo=T}
broom::glance(myLinearModel)
```

---

# Our first model

We do need some sense as to how well this model fit the data

.pull-left[
```{r Rintro-39, echo=T, eval=F}
# install.packages('ggfortify')
library(ggfortify)
autoplot(myLinearModel)
```
]
.pull-right[
```{r Rintro-40, echo=F, eval=T, fig.height=6}
# install.packages('ggfortify')
library(ggplot2)
library(ggfortify)
autoplot(myLinearModel)

```
]


---

## Displaying model results

Let's start with this model:

```{r Rintro-41}
myModel <- lm(log10(enterococci) ~ rainfall + temperature + season_name + factor(year), data = beaches)
broom::tidy(myModel)
```

---

# Displaying model results

Let's start with this model:

```{r Rintro-42}
plt_data <- broom::tidy(myModel)
plt_data <- plt_data %>% filter(term != '(Intercept)') %>% #<<
  mutate(term = str_replace(term, 'season_name','')) #<<
plt_data
```

---

# Displaying model results

Let's start with this model:

```{r Rintro-43}
plt_data <- broom::tidy(myModel)
plt_data <- plt_data %>% filter(term != '(Intercept)') %>% 
  mutate(term = str_replace(term, 'season_name','')) %>% 
  mutate(term = str_replace(term, 'factor\\(year\\)','')) # Brackets are "escaped" using \\ #<< 
plt_data
```

---

# Displaying model results

Let's start with this model:

.pull-left[
```{r c1, eval=F}
plt_data <- broom::tidy(myModel)
plt_data %>% filter(term != '(Intercept)') %>% 
  mutate(term = str_replace(term, 'season_name','')) %>% 
  mutate(term = str_replace(term, 'factor\\(year\\)','')) %>% # Brackets need to be "escaped" using \\
  ggplot(aes(x = term, y = estimate, 
             ymin = estimate - 2 * std.error, 
             ymax = estimate + 2 * std.error))+
    geom_pointrange()

```
]
.pull-right[
```{r Rintro-44, ref.label='c1', echo=F, eval=T, fig.height=6}

```

]

---

# Displaying model results

Let's start with this model:

.pull-left[
```{r c2, eval=F}
plt_data <- broom::tidy(myModel)
plt_data %>% filter(term != '(Intercept)') %>% 
  mutate(term = str_replace(term, 'season_name','')) %>% 
  mutate(term = str_replace(term, 'factor\\(year\\)','')) %>% # Brackets need to be "escaped" using \\
  ggplot(aes(x = term, y = estimate, 
             ymin = estimate - 2 * std.error, 
             ymax = estimate + 2 * std.error))+
    geom_pointrange() + 
    geom_hline(yintercept = 0, linetype=2) + 
    theme_bw() + 
    coord_flip()
  

```
]
.pull-right[
```{r Rintro-45, ref.label='c2', echo=F, eval=T, fig.height=6}

```

]
---

# Displaying model results

Let's start with this model:

.pull-left[
```{r Rintro-46, eval=F}
plt_data <- broom::tidy(myModel)
plt_data %>% filter(term != '(Intercept)') %>% 
  mutate(term = str_replace(term, 'season_name','')) %>% 
  mutate(term = str_replace(term, 'factor\\(year\\)','')) %>% # Brackets need to be "escaped" using \\
  ggplot(aes(x = term, y = estimate, 
             ymin = estimate - 2 * std.error, 
             ymax = estimate + 2 * std.error))+
    geom_pointrange() + 
    geom_hline(yintercept = 0, linetype=2) + 
    theme_bw() + 
    coord_flip()

ggsave('results.png') # ggsave knows format from file
  

```

You can also save the graph from the RStudio Plots pane, but coding it using `ggsave` is more reproducible

If you need to get a high-definition TIFF, your best bet is to save your graph as a PDF and then convert it using Acrobat or other scripts (ask me if interested). The TIFF printer in R only creates 72 DPI TIFF files
]
.pull-right[
```{r Rintro-47, ref.label='c2', echo=F, eval=T, fig.height=6}

```

]

---

## Showing group differences ("Figure 1")

The package `ggpubr`, which extends `ggplot2`, makes this very easy. It provides the function `stat_compare_means`

---

## Showing group differences ("Figure 1")

.pull-left[
```{r d1, eval=F}
library(ggpubr)
theme_viz <- function(){ #<<
  theme_bw() + #<<
    theme(axis.title = element_text(size=16),#<<
          axis.text = element_text(size=14),#<<
          text = element_text(size = 14)) #<<
} #<<
ggplot(
  data=beaches, 
  mapping= aes(x = season_name, y = log10(enterococci), color = season_name)) +
  geom_boxplot()+geom_jitter()+
  labs(x = 'Season', y = expression(paste('log'['10'],'(enterococci)')), color='Season') +
  theme_viz()
```
]
.pull-right[
```{r Rintro-48, ref.label='d1', eval=T, echo=F }

```

]

---

## Showing group differences ("Figure 1")

.pull-left[
```{r d2, eval=F}
library(ggpubr)
plt <- ggplot(
  data=beaches, 
  mapping= aes(x = season_name, y = log10(enterococci), color = season_name)) +
  geom_boxplot() + 
  geom_jitter(width=0.1) +
  labs(x = 'Season', y = expression(paste('log'['10'],'(enterococci)')), color='Season')+
  theme_viz()
my_comparisons <- list(c('Autumn','Spring'),
                       c('Spring','Summer'),
                       c('Summer','Winter'),
                       c('Spring','Winter'))
plt + stat_compare_means() #<<
```
]
.pull-right[
```{r Rintro-49, ref.label='d2', eval=T, echo=F }

```

]

---

## Showing group differences ("Figure 1")

.pull-left[
```{r d3, eval=F}
library(ggpubr)
plt <- ggplot(
  data=beaches, 
  mapping= aes(x = season_name, y = log10(enterococci), color = season_name)) +
  geom_boxplot() + 
  geom_jitter(width=0.1)+
    labs(x = 'Season', y = expression(paste('log'['10'],'(enterococci)')), 
         color='Season') +
  theme_viz()
my_comparisons <- list(c('Autumn','Spring'),
                       c('Spring','Summer'),
                       c('Summer','Winter'),
                       c('Spring','Winter'))
plt + stat_compare_means() +
  stat_compare_means(comparisons = my_comparisons) #<<
```
]
.pull-right[
```{r Rintro-50, ref.label='d3', eval=T, echo=F }

```

]

---

## Showing group differences ("Figure 1")

.pull-left[
```{r d5, eval=F}
library(ggpubr)
plt <- ggplot(
  data=beaches, 
  mapping= aes(x = season_name, y = log10(enterococci), color = season_name)) +
  geom_boxplot() + 
  geom_jitter(width=0.1)+
  labs(x = 'Season', y = expression(paste('log'['10'],'(enterococci)')), color='Season')+
  theme_viz()
my_comparisons <- list(c('Autumn','Spring'),
                       c('Spring','Summer'),
                       c('Summer','Winter'),
                       c('Spring','Winter'))
plt + stat_compare_means(label.y = 6) + 
  stat_compare_means(comparisons = my_comparisons) 
```
]
.pull-right[
```{r Rintro-51, ref.label='d5', eval=T, echo=F }

```

]

---

## Manipulating data for plotting

We would like to get density plots of all the variables

.pull-left[
```{r Rintro-52}
dat_spine <- rio::import('data/Dataset_spine.csv', check.names=T)
head(dat_spine)
```
]
.pull-right[
Facets only work by grouping on a variable. Here we have data in several columns
]

---

## Manipulating data for plotting

We would like to get density plots of all the variables. 

.pull-left[
```{r Rintro-53}
dat_spine %>% 
  tidyr::gather(variable, value, everything())
```
]
.pull-right[
The `gather` function turns this wide dataset to a long dataset, stacking all the variables on top of each other
]

---

## Manipulating data for plotting

We would like to get density plots of all the variables. 

.pull-left[
```{r Rintro-54, fig.height=3}
dat_spine %>% 
  select(Pelvic.incidence:Sacral.slope) %>% 
  tidyr::gather(variable, value) %>% 
  ggplot(aes(x = value)) + 
  geom_density() +
  facet_wrap(~variable) +
  labs(x = '')
```
]
.pull-right[
This is one of my most used tricks for getting facetted plots from wide data
]

---

## Re-ordering factors

```{r e1, fig.height=4}
beaches %>% 
  ggplot(aes(x = season_name, y = temperature)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::unit_format(unit = "\u00B0C")) +
  labs(x = 'Season', y = 'Temperature') +
  theme_bw() + 
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```

---

## Re-ordering factors

```{r e2, fig.height=4}
beaches %>% 
  mutate(season_name = fct_relevel(season_name, 'Autumn','Winter','Spring','Summer')) %>% #<<
  ggplot(aes(x = season_name, y = temperature)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::unit_format(unit = "\u00B0C")) +
  labs(x = 'Season', y = 'Temperature') +
  theme_bw() + 
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14))
```
