---
title: "A sampling of bioinformatics using R"
author: Abhijit Dasgupta, PhD
date: Spring 2019

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F, comment = '#> ', cache=FALSE, fig.height=6, cache=F)
library(tidyverse)
```

layout: true

<div class="my-header">
<span>BIOF 439, Spring 2019</span></div>

---

## Expectations

1. We will explore
    - the basics of using Bioconductor packages for bioinformatics
    - visualizations useful for bioinformatics
    - ways to annotate our graphics
1. We will not explore
    - how to do bioinformatics using R
    - bioinformatic workflows
    - data cleaning, modeling, analytics

This area is too broad and interests too varied to cover this with any sort of breadth or depth. 

---

## A very simple example

```{r}
fix_names <- function(d) d %>% set_names(make.names(names(.)))
clinical <- rio::import('data/BreastCancer_Clinical.xlsx') %>% fix_names()
proteome <- rio::import('data/BreastCancer_Expression.xlsx') %>% fix_names()
final_data <- clinical %>% 
    inner_join(proteome, by = c('Complete.TCGA.ID' = 'TCGA_ID')) %>% 
    filter(Gender == 'FEMALE') %>% 
    select(Complete.TCGA.ID, Age.at.Initial.Pathologic.Diagnosis, ER.Status, starts_with("NP"))
head(final_data)
```

---

## A very simple example

```{r}
results <- final_data %>% 
    summarise_at(vars(starts_with('NP')), #<<
                 ~wilcox.test(. ~ ER.Status)$p.value)#<<
results
```

`.` is the placeholder for what's specified inside the `vars()`. 

--

This isn't in the right format for me to plot

---

## A very simple example

.pull-left[
```{r}
results %>% tidyr::gather(protein, pvalue)
```
]

---

## A very simple example

.pull-left[
```{r a1, eval = F}
theme_439 <- theme_bw() + 
    theme(axis.title = element_text(size=16),
          axis.text = element_text(size=8))


results %>% tidyr::gather(protein, pvalue) %>% 
    ggplot(aes(x = protein, y = pvalue)) + 
      geom_point() +
    theme_439
```
]
.pull-right[
```{r, ref.label='a1', echo=F, eval = T}

```

]

---

## A very simple example

.pull-left[
```{r a2, eval = F}
results %>% tidyr::gather(protein, pvalue) %>% 
    ggplot(aes(x = protein, y = pvalue)) + 
      geom_point() +
    geom_segment(aes(x = protein, xend = protein, #<<
                     y = 0, yend = pvalue))+ #<<
    theme_439 
```
]
.pull-right[
```{r, ref.label='a2', echo=F, eval = T}

```

]

---

## A very simple example

.pull-left[
```{r a3, eval = F}
results %>% tidyr::gather(protein, pvalue) %>% 
    ggplot(aes(x = protein, y = pvalue)) + 
      geom_point() +
    geom_segment(aes(x = protein, xend = protein, 
                     y = 0, yend = pvalue))+
    geom_hline(yintercept = 0.05, linetype=2)+ #<<
    theme_439 
```
]
.pull-right[
```{r, ref.label='a3', echo=F, eval = T}

```

]

???

This is basically the beginnings of a Manhattan plot

---

## Manhattan plot

A Manhattan plot is used to visualize a set of p-values from unit-based tests

It plots the negative log p-value at each unit

.pull-left[
```{r a4, eval=F}
results %>% tidyr::gather(protein, pval) %>% 
    ggplot(aes(x = protein, y = -log10(pval))) +
    geom_point() + 
    geom_segment(aes(xend = protein, yend = 0))+
    geom_hline(yintercept = 8, linetype=2)+
    labs(x = 'Protein',
         y = expression(log[10](p-value))) +
    theme_439
```
]
.pull-right[
```{r, ref.label='a4', echo=F, eval=T, fig.height=4}

```

]

---

## Manhattan plot

There is a specialized package for doing Manhattan plots and quantile plots for GWAS data

This package is meant to work with PLINK output, but the function is generic

.pull-left[
```{r b1, eval = F}
library(qqman) #<<
manhattan(gwasResults)
```
]
.pull-right[
```{r, ref.label='b1', eval=T, echo=F, fig.height=6}

```
]

---

## Manhattan plot

.pull-left[
```{r b2, eval = F}
library(qqman) #<<
manhattan(gwasResults, 
          annotatePval = 1e-6, 
          annotateTop=F)
```
]
.pull-right[
```{r, ref.label='b2', eval=T, echo=F}

```
]

???

Look at the help for manhattan

---

## ggplot tours Manhattan

### Data prep

.pull-left[
```{r c1,eval=F}
head(gwasResults)
```
]
.pull-right[
```{r, ref.label='c1', eval=T, echo=F}

```
]

---

## ggplot tours Manhattan

### Data prep

.pull-left[
```{r c2, eval =F}
plt_x_position <- gwasResults %>% 
    group_by(CHR) %>% 
    summarize(chr_len = max(BP)) %>% 
    mutate(tot = cumsum(chr_len) - chr_len)
head(plt_x_position)
```

]
.pull-right[
```{r, ref.label='c2', eval=T, echo=F}

```

]

---

## ggplot tours Manhattan

### Data prep

.pull-left[
```{r c3, eval = F}
plt_dat <- gwasResults %>% 
    left_join(plt_x_position %>% select(-chr_len), 
              by = c('CHR'='CHR')) %>% 
    arrange(CHR, BP) %>% 
    mutate(BPcumul = BP + tot)
tail(plt_dat)
```

]
.pull-right[
```{r, ref.label='c3', eval=T, echo =F}

```
]

---

## ggplot tours Manhattan

### Data for plotting x-axis labels

.pull-left[
```{r c4, eval = F}
axisdf <- plt_dat %>% 
    group_by(CHR) %>% 
    summarize(center = (max(BPcumul)+min(BPcumul))/2)
axisdf
```

]
.pull-right[
```{r, ref.label='c4', eval=T, echo =F}

```

]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c5, eval = F}
ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point() 
```

]
.pull-right[
```{r, ref.label='c5', eval=T, echo =F}

```

]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c6, eval = F}
ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point(aes(color = as.factor(CHR)), #<<
               alpha = 0.8, size=1.3) #<<
```

]
.pull-right[
```{r, ref.label='c6', eval=T, echo =F}

```

]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c7, eval = F}
ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point(aes(color = as.factor(CHR)), 
               alpha = 0.8, size=1.3) +
    scale_color_manual(
        values = rep(c('grey','skyblue'), 22))
```

]
.pull-right[
```{r, ref.label='c7', eval=T, echo =F}

```

]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c8, eval = F}
ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point(aes(color = as.factor(CHR)), 
               alpha = 0.8, size=1.3) +
    scale_color_manual(
        values = rep(c('grey','skyblue'), 22))+
    scale_x_continuous(#<<
        name = 'Chromosome',#<<
        breaks = axisdf$center,#<<
        labels = axisdf$CHR#<<
    )#<<
```

]
.pull-right[
```{r, ref.label='c8', eval=T, echo =F}

```

]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c9, eval = F}
ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point(aes(color = as.factor(CHR)), 
               alpha = 0.8, size=1.3) +
    scale_color_manual(
        values = rep(c('grey','skyblue'), 22))+
    scale_x_continuous(
        name = 'Chromosome',
        breaks = axisdf$center,
        labels = axisdf$CHR
    ) +
    scale_y_continuous(expand = c(0,0)) + #<<
    theme_bw()
```

]
.pull-right[
```{r, ref.label='c9', eval=T, echo =F}

```

]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c10, eval = F}
ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point(aes(color = as.factor(CHR)), 
               alpha = 0.8, size=1.3) +
    scale_color_manual(
        values = rep(c('grey','skyblue'), 22))+
    scale_x_continuous(
        name = 'Chromosome',
        breaks = axisdf$center,
        labels = axisdf$CHR
    ) +
    scale_y_continuous(expand = c(0,0)) + #<<
    theme_bw() + 
    geom_point(data = plt_dat %>% filter(P < 1e-5),#<<
               color = 'orange', 
               size=2)
```

]
.pull-right[
```{r, ref.label='c10', eval=T, echo =F}

```
]

---

## ggplot tours Manhattan

### Plotting

.pull-left[
```{r c11, eval = F}
(plt_manhattan <- ggplot(plt_dat, 
       aes(x = BPcumul, y = -log10(P)))+
    geom_point(aes(color = as.factor(CHR)), 
               alpha = 0.8, size=1.3) +
    scale_color_manual(
        values = rep(c('grey','skyblue'), 22))+
    scale_x_continuous(
        name = 'Chromosome',
        breaks = axisdf$center,
        labels = axisdf$CHR
    ) +
    scale_y_continuous(expand = c(0,0)) + 
    theme_bw() + 
    geom_point(data = plt_dat %>% filter(P < 1e-5),#<<
               color = 'orange', 
               size=2)+
    geom_hline(yintercept = -log10(5e-8), linetype=2)+
    theme(legend.position='none',
          panel.grid.major.x=element_blank(),
          panel.grid.minor.x = element_blank())
)
```

]
.pull-right[
```{r, ref.label='c11', eval=T, echo =F}

```
]

---

## ggplot tours Manhattan

### Annotation

.pull-left[
```{r c12, eval = F}
library(ggrepel) #<<
plt_manhattan +
    geom_label_repel(
        data = plt_dat %>% filter(P < 1e-5),
        aes(label = SNP),
        size = 2)
```

]
.pull-right[
```{r, ref.label='c12', eval=T, echo =F}

```
]

---

## Circular Manhattan plots

.pull-left[
```{r d1, eval =F}
library(CMplot)
CMplot(gwasResults, plot.type = 'c',
       r = 1.6,
       cir.legend = T, 
       outward=T,
       cir.legend.col='black',
       cir.chr.h=.1,
       chr.den.col='orange',
       file.output=F)
```

]
.pull-right[
```{r, ref.label='d1', eval=T, echo=F, message=F}

```

]


---
class: middle,center

# Heatmaps

---

## Let us count the ways

There are several ways of doing heatmaps in R:

+ [http://sebastianraschka.com/Articles/heatmaps_in_r.html](http://sebastianraschka.com/Articles/heatmaps_in_r.html)
+ [https://plot.ly/r/heatmaps/](https://plot.ly/r/heatmaps/)
+ [http://moderndata.plot.ly/interactive-heat-maps-for-r/](http://moderndata.plot.ly/interactive-heat-maps-for-r/)
+ [http://www.siliconcreek.net/r/simple-heatmap-in-r-with-ggplot2](http://www.siliconcreek.net/r/simple-heatmap-in-r-with-ggplot2)
+ [https://rud.is/b/2016/02/14/making-faceted-heatmaps-with-ggplot2/](https://rud.is/b/2016/02/14/making-faceted-heatmaps-with-ggplot2/)

---

## Some example data

```{r}
library(Biobase)
data(sample.ExpressionSet)
exdat <-  sample.ExpressionSet
library(limma)
design1 <- model.matrix(~type, data=pData(exdat))
lm1 <- lmFit(exprs(exdat), design1)
lm1 <- eBayes(lm1) # compute linear model for each probeset
geneID <- rownames(topTable(lm1, coef=2, num=100, adjust='none',p.value=0.05))
exdat2 <- exdat[geneID,] # Keep features with p-values < 0.05

head(exdat2)
```

---

## Using Heatplus

.pull-left[
```{r e1, eval=F}
# BiocManager::install('Heatplus')
library(Heatplus)
reg1 <- regHeatmap(exprs(exdat2), legend=2, col=heat.colors, 
                   breaks=-3:3)
plot(reg1)
```

]
.pull-right[
```{r, ref.label='e1', eval=T, echo=F}

```
]

---

## Using Heatplus

.pull-left[
```{r e2, echo=T, eval=F}
corrdist <- function(x) as.dist(1-cor(t(x)))
hclust.avl <- function(x) hclust(x, method='average')
reg2 <- regHeatmap(exprs(exdat2), legend=2, 
                   col=heat.colors,
                   breaks=-3:3, 
                   dendrogram = 
                       list(clustfun=hclust.avl, 
                            distfun=corrdist))
plot(reg2)
```
]
.pull-right[
```{r, ref.label='e2', eval=T, echo=F}

```
]

---

## Using Heatplus

### Adding annotations

.pull-left[
```{r e3, echo=T, eval=F}
ann1 <- annHeatmap(exprs(exdat2), 
                   ann=pData(exdat2), #<<
                   col = heat.colors)
plot(ann1)
```
]
.pull-right[
```{r, ref.label='e3', echo=F, eval=T}

```

]

---

## Using Heatplus

### Adding annotations

.pull-left[
```{r e4, echo=T, eval=F, fig.width=7, fig.height=6}
ann2 <- annHeatmap(exprs(exdat2), 
                   ann=pData(exdat2), 
                   col = heat.colors,
        cluster = 
            list(cuth=7500,
                 label=c('Control-like','Case-like')))
plot(ann2)
```
]
.pull-right[
```{r, ref.label='e4', eval=T, echo=F}

```

]

---
class:middle, center

# Playing with Seurat

---

## Example data

```{r}
library(Seurat)
# pbmc.data <- Read10X(data.dir='data/hg19/')
# pbmc <- CreateSeuratObject(counts = pbmc.data, project='pbmc3k', min.cells=3, min.features=200)
pbmc <- readRDS('data/pbmc.rds')
pbmc
names(pbmc)
slotNames(pbmc)
```

---

## Adding QC metrics and plotting

We'll calculate mitochondrial QC metrics (percentage counts originating from mitochondrial genes)

```{r}
pbmc[['percent.mt']] <- PercentageFeatureSet(pbmc, pattern = '^MT-')
head(pbmc@meta.data) #<<
```

---

## Visualizing metrics

.pull-left[
```{r f1, eval=F}
# plt <- VlnPlot(object = pbmc, 
#   features = c('nFeature_RNA',
#                'nCount_RNA',
#                'percent.mt'))

plot_data <- pbmc@meta.data %>%
  tidyr::gather(variable, value, -orig.ident)

ggplot(plot_data, aes(orig.ident, value)) +
  geom_violin(fill = 'red') +
  geom_jitter(width=0.5, alpha = 0.1) +
  facet_wrap(~variable, nrow = 1, 
             scales = 'free_y') + #<<
  labs(x = 'Identity',y = 'Expression Level') +
  theme_classic()
```

]
.pull-right[
```{r, ref.label='f1', eval=T, echo=F}

```

]

---

## Visualizing feature-feature relationships

.pull-left[
```{r g1, eval=F}
plot1 <- FeatureScatter(object = pbmc, 
                        feature1 = "nCount_RNA", 
                        feature2 = "percent.mt")
plot2 <- FeatureScatter(object = pbmc, 
                        feature1 = "nCount_RNA", 
                        feature2 = "nFeature_RNA")
CombinePlots(plots = list(plot1, plot2))
```
]
.pull-right[
```{r, ref.label='g1', eval=T, echo=F}

```

]

---

## Visualizing feature-feature relationships

.pull-left[
```{r g2, eval=F}
cormatrix <- cor(pbmc@meta.data %>% select(-orig.ident))
plt1 <-
  ggplot(pbmc@meta.data, 
         aes(x = nCount_RNA, 
             y = percent.mt, 
             group = orig.ident, 
             color = orig.ident)) +
  geom_point() +
    theme_classic() +
    labs(color = 'Identity',
         title=as.character(round(cormatrix['nCount_RNA','percent.mt'],2)))+
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

plt2 <-
  ggplot(pbmc@meta.data, 
         aes(x = nCount_RNA, 
             y = nFeature_RNA, 
             group = orig.ident, 
             color = orig.ident)) +
  geom_point() +
  theme_classic() +
  labs(color = 'Identity',
       title=as.character(round(cormatrix['nCount_RNA','nFeature_RNA'],2)))+
  theme(plot.title = element_text(face = 'bold', hjust = 0.5))

ggpubr::ggarrange(plt1, plt2, nrow = 1, ncol=2)

```
]
.pull-right[
```{r, ref.label='g2', eval=T, echo=F}

```

]

---

## Feature selection

.pull-left[
```{r g3, eval=F}
pbmc <- subset(x = pbmc, 
    subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
pbmc <- NormalizeData(object = pbmc, 
                      normalization.method = "LogNormalize", 
                      scale.factor = 10000)
# This is stored in pbmc[['RNA']]@meta.features

pbmc <- FindVariableFeatures(object = pbmc, 
                             selection.method = "vst", 
                             nfeatures = 2000)


# Identify the 10 most highly variable genes
top10 <- head(x = VariableFeatures(object = pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(object = pbmc)
plot1

```

]
.pull-right[
```{r, ref.label='g3', eval=T, echo=F}

```

]

---

## Feature selection

.pull-left[
```{r g4, eval=F}
plt_data <- pbmc[['RNA']]@meta.features %>% 
    rownames_to_column(var='id')
topvars <- pbmc[['RNA']]@var.features
plt_data <- plt_data %>% 
    mutate(indic = ifelse(id %in% topvars, 
                          'Variable count',
                          'Non-variable count'))
bl <- plt_data %>% 
    count(indic) %>% 
    glue::glue_data("{indic}: {n}")
names(bl) <- c('Non-variable count','Variable count')
plt_data <- plt_data %>%
  mutate(indic = bl[indic])
plt11 <- ggplot(plt_data, 
                aes(x = mean, 
                    y = variance.standardized, 
                    color = indic)) +
  geom_point() +
  scale_x_log10() +
  scale_color_manual(values = c('black','red')) +
  labs(x = 'Average Expression', y = 'Standardized Variance', color = '')+
  theme_classic()
plt11
```

]
.pull-right[
```{r, ref.label='g4', eval=T, echo=F}

```

]

---

## Feature selection

.pull-left[
```{r g5, eval=F}
# plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plt12 <- plt11 + ggrepel::geom_text_repel(data = plt_data %>% filter(id %in% top10),
                                          aes(label = id),
                                          color = 'black')
plt12
```

]
.pull-right[
```{r, ref.label='g5', eval=T, echo=F}

```
]
---

## There's a lot more

We'll stop our sampling here. 

+ Many Bioconductor packages do use ggplot, however some use base graphics
    - Faster
+ Key is to find where the data is stored, and use that to create visualizations
+ Bioconductor tends to create
    - One monolithic object
    - Containing different information in slots 
    - combined by lists
+ `slotNames` and `names` are your friends
