---
title: "R Project"
author: "Lauren"
date: "5/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Final project

- Look at the relationship between playoff overtime goals and play off win percentage for NHL teams 


##Library list
```{r library, include=F}
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(plotly)) install.packages("plotly")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(readxl)
library(plyr)
library(plotly)
```

##Data import/set up
```{r data, eval=T}
nhl_rec <- read_xls("nhl_r.xls", sheet = 1)
nhl_rec_team <- as_tibble(nhl_rec)
nhl_ot <- read_xls("nhl_r.xls", sheet = 2)
nhl_ot_team <- count(nhl_ot, "Team")
names(nhl_rec)[1] = 'TeamID'
nhl_data1 <- nhl_rec %>% left_join(nhl_ot_team, by = c('TeamID'='Team'))
nhl_data1[is.na(nhl_data1)] <- 0
head(nhl_data1)
nhl_data2 <- nhl_rec %>% left_join(nhl_ot, by = c("TeamID" = "Team"))
  nhl_data2$Date <- as.Date(nhl_data2$Date, format = "%Y-%m-%d")
head(nhl_data2)
```

##Rough win% and OT goal freq plot
- Graph of total playoff OT goals and overall win percentage 
- this graph fails to take into account that some teams have been in the league for far longer than others
```{r plot1, eval=T}
mod1 <- lm(freq ~ Pct., data = nhl_data1)
r2 <- broom::glance(mod1) %>% pull(r.squared) %>% 
  round(., 2)

plt1 <- ggplot(data = nhl_data1, aes(x = freq, y = Pct.)) + # Check your variable names
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs( title = "Total OT goals and Win Percentage",
        subtitle = "All time playoff OT goals and playoff win percentage",
        x = "Total OT Goals",
        y = "Win %") + 
  annotate('text',
           x = 60, y = 0.07,
           label=glue::glue("R^2 == {r}",r=r2),
           size=6,
           parse=T) +
  theme_minimal()
plt1
```

##OT goal freq following 1991 
- many of the big changes to the composition of the NHL were made prior to 1991, so I decided to start looking into the goals scored only following 1991
- still not the best way to do it
```{r plot2, eval = T}
library(lubridate)

nhl <- nhl_data2 %>% filter(Date >= ("1991-01-01")) %>% count("TeamID")
names(nhl)[2] = "OTg_post_1991"
nhl

plt2 <- ggplot(data = nhl, mapping = aes(x = reorder(TeamID, OTg_post_1991), y = OTg_post_1991)) +
  geom_bar(stat = "Identity") +
  labs( x = "Team", 
        y = "OT goal count",
        title = "OT Goal Count by Team",
        subtitle = "Goals 1991 - Present") + 
  theme_minimal()+
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  coord_flip()
  
plt2
```

## Calculating goals per game average
- instead of looking into certain dates I decided to calculate the average number of goals per game, this would better account for teams who have scored more playoff OT goals due to more playoff appearances 
```{r plt3, eval = T}

nhl_data3 <- cbind(nhl_data1, nhl_data1$W+nhl_data1$L)
names(nhl_data3)[7] = 'Total games'

nhl_data4 <- cbind(nhl_data3, nhl_data3$freq/nhl_data3$`Total games`)
names(nhl_data4)[8] = "Gpg"

nhl_data5 <- nhl_data4 %>% 
  mutate_if(is.numeric, round, digits = 2)

head(nhl_data5)

```

## Putting it together
```{r plt4, eval = T}
mod2 <- lm(Gpg ~ Pct., data = nhl_data5)
r2_ <- broom::glance(mod2) %>% pull(r.squared) %>% 
  round(., 2)

nhl_theme <- theme_minimal() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        axis.ticks = element_blank())

plt4 <- ggplot(nhl_data5, aes(x = Gpg, y = Pct.)) +
  geom_point(color = "blue") +
  geom_smooth(method = lm, se = FALSE, color = "red")  + 
  annotate('text',
           x = 0.5, y = 0.07,
           label=glue::glue("R^2 == {r}", r=r2_),
           size=6,
           parse=T) +  
      labs(x = "OT Goals Per Game",
      y = "Win Percentage",
      title = "Playoff OT Goals per Game and Win Percentage") +
  nhl_theme
      
plt4

```

## Making it interactive 
- I also wanted to compare the functionality of using ggplotly and plotly to plot the same data 
```{r plt5, eval = T}
library(plotly)
ggplotly(plt4)

fit <- lm(Pct. ~ Gpg , data = nhl_data5)
plt5 <- plot_ly(data = nhl_data5, x = nhl_data5$Gpg, y = nhl_data5$Pct., type = "scatter", 
        mode = "markers", text = nhl_data5$TeamID) %>% 
  add_lines(x = ~Gpg, y = fitted(fit)) %>% 
  layout(title = "OT Goals per Game vs Win Percentage", showlegend = F, 
         xaxis = list(title = "Goals per Game", showgrid = F),
         yaxis = list(title = "Win %", showgrid = F))
 plt5
```