---
title: "Homework 1"
author: "BIOF 439, Spring 2019"
date: "Due date: Tuesday, April 9 at midnight"
output: html_document
---

-----

### Readings

R for Data Science [https://r4ds.had.co.nz](https://r4ds.had.co.nz)

- Chapter 4 (Workflow: Basics)
- Chapter 27 (R Markdown)

### Hands-on practice

On RStudio Cloud, you will have a new project called Homework 1, which has a R Markdown document called `hw1.Rmd`. Please open it, follow the instructions within it, answer
the questions, and then compile to HTML (see [instructions](instructions.html))

-----

## Exercises

### Question 1

Run the following code, either in your console or from the R Markdown document
```{r, eval=F}
living_wage <- 15
l1ving_wage
```

Does this code work? Why or why not? How would you correct it? Put the corrected code in a R code chunk in your answer, with option `eval = T` to show that the code works

__Answer:__





### Question 2

Complete the code below by repalcing the `___`, following the instructions in the comments (lines starting with `#`). Change `eval = F` to `eval = T`, so that the answers show in the HTML file. You can learn more about the dataset by typing `?mtcars` in the console.

```{r warmup, message = F, warning = F, eval = F}
library(tidyverse)

data(mtcars)

# Show me the first 3 rows of this dataset

___(mtcars, n = ___)

# What is the average mpg of all the cars?

___(mtcars$___)


# Find the average mpg for 4 cylinder cars

___(mtcars$mpg[mtcars$___ == 4])


# Find the average mpg for 4, 6 and 8 cylinder cars, in one statement

mtcars %>% 
  group_by(___) %>% 
  summarize(avg_mpg = mean(___))


# Convert mpg to km/l (1 mpg = 1.6 km/g = 0.42 km / l) and compute avg mpg & km/l by cylinder

mtcars %>% 
  mutate( kml = mpg * ___) %>% 
  group_by(___) %>% 
  summarize(avg_US = mean(___), avg_metric = ____(____))
  
```

### Question 3A first pass at graphs

There is some template ggplot2 code below. Once again fill in the blanks and change `eval = F` to `eval = T` to run the code during knitting.

```{r graphs1, eval = F}
library(ggplot2)

# Draw a scatter plot showing the relationship between displacement and mpg

ggplot(mtcars, aes(x = ___, y = ___)) +
  geom_point()
```

```{r graphs2, eval = F}

# Draw a scatterplot like above, but color points based on cylinders
# 

ggplot(mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(color = ___))
```

_Extra credit:_  
Is this quite right? Can you fix it by changing the data type of `cyl`?


