---
title: "Data Viz Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    source_code: embed
    
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(plotly)
knitr::opts_chunk$set(message = FALSE)
```

Original Data {vertical_layout=scroll data-icon="fa-archive"}
===================================== 

Row {data-height=10000}
------------------------------------------------------------------------------

### A look at the data

```{r raw data}
EchoC_data <- read_excel("Pfeifer Echo Edit for R.xlsx", sheet = "Echo_for_R")

#AD EchoC_data <- read_excel('Pfeifer Echo Edit for R.xlsx', sheet = "Echo_for_R") %>% 
#AD  mutate(Genotype = as.factor(Genotype))
head(EchoC_data)
summary(EchoC_data)

#AD Look at tableone::createTableOne for some better options than `summary`

names(EchoC_data)

```

Column {data-height=10000}
------------------------------------------------------------------------------

### Running some statistics

```{r stats}

stats_1<-lapply(EchoC_data[,4:24], function(x) t.test(x ~ EchoC_data$Genotype, var.equal = TRUE))
stats_1

#AD A more compact representation might be:
#AD 
#AD stats_1_AD <- EchoC_data %>% 
#AD   gather(variable, value, -(1:3)) %>% 
#AD   nest(-variable) %>% 
#AD   mutate(mods = map(data, ~t.test(value ~ Genotype, data = ., var.equal = T))) %>% 
#AD   mutate(result = map(mods, broom::tidy)) %>% 
#AD   select(variable, result) %>% 
#AD   unnest()
#AD 
#AD This idea is expanded in https://r4ds.had.co.nz/many-models.html
#AD 
#AD If I start from where you left off, this gets to the same place:
#AD stats_1 %>% map(broom::tidy) %>% bind_rows(.id = 'variable') 
#AD 
#AD Generally using broom::tidy is very useful for compact presentation of results
```


Visualizations {data-orientation=columns data-icon="fa-chart-bar"}
===================================== 
Row {.tabset data-height=400}
------------------------------------------------------------------------------

### Early violin plots 

```{r Early Data}
LV_Sys_Vol_Sept <- ggplot(EchoC_data, aes(x = Genotype, y = LV_Vol_s)) + 
  geom_violin() + 
  theme_bw() + 
  ylab("Left Ventricular Systolic Volume (uL)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant"), label = c("Wild Type", "Mutant")) +
  ggtitle("September 2018 Systolic Volume Variation") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_summary(fun.y = mean, geom="point", shape = 18, size  = 3) + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue")
LV_Sys_Vol_Sept

LV_Diameter_Sept <- ggplot(EchoC_data, aes(x = Genotype, y = Dia_s)) + 
  geom_violin() + 
  theme_bw() + 
  ylab("Left Ventricular Diameter (mm)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant"), label = c("Wild Type", "Mutant")) +
  ggtitle("September 2018 Diameter Variation During Systole") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_summary(fun.y = mean, geom="point", shape = 18, size  = 3) + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue")
LV_Diameter_Sept

#AD A suggestion here is to bundle code for ggplot into functionality. So 
#AD geoms and stats first, then scales, labels and themes. So I would prefer to write:
#AD 
#AD ggplot(EchoC_data, aes(x = Genotype, y = Dia_s)) +
#AD   geom_violin() + 
#AD   stat_summary(fun.y = mean, geom = "point", shape = 18, size = 3) + 
#AD   stat_summary(fun.y = median, geom = "point", shape = 3, size = 3, color = 'blue') +
#AD   scale_x_discrete(limits = c("Wild_Type", "Mutant"), label = c("Wild Type", "Mutant")) +
#AD   ylab('Left Ventricular Diameter (mm)') + 
#AD   ggtitle('September 2018 Diameter Variation During Systole') +
#AD   theme_bw() +
#AD   theme(plot.title = element_text (hjust = 0.5))
#AD
#AD This helps me cognitively, since with your style I find myself searching for
#AD elements and so find it harder to understand and debug
```

Row {.tabset data-height=400}
------------------------------------------------------------------------------

### Poster violin plots 

```{r poster graphs}

jn_theme <- function(){
  theme_bw() +
    theme(axis.text = element_text(size = 14, color = "Black"),
          axis.title = element_text(size = 16),
          panel.grid.minor = element_blank(),
          strip.text = element_text(size=14),
          strip.background = element_blank(),
          plot.title = element_text(size = 20, hjust = 0.5),
          panel.grid.major.x = element_blank())
}


LV_Sys_Vol_15 <- ggplot(data = EchoC_data[EchoC_data$Group == "September",], aes(x = Genotype, y = LV_Vol_s)) + 
  geom_violin(scale = "area", adjust = 1, aes(color = Genotype), show.legend = FALSE) + 
  scale_color_manual(values = c("Wild_Type"="Black", "Mutant"="Red")) +
  geom_point(size = 0.5, height = 0, width = 0.05) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Left Ventricular Systolic Volume (uL)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  stat_summary(fun.y = mean, geom = "point", shape = 5, size  = 3, color = "#9A2617") + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue") +
  jn_theme() +
  theme(axis.title.x = element_blank())
LV_Sys_Vol_15

#Dont need this for poster
LV_Diam_systole_15 <- ggplot(data = EchoC_data[EchoC_data$Group == "September",], aes(x = Genotype, y = Dia_s)) + 
  geom_violin(scale = "area", adjust = 1, aes(color = Genotype), show.legend = FALSE) + 
  scale_color_manual(values = c("Wild_Type"="Black", "Mutant"="Red")) +
  geom_point(size = 0.5, height = 0, width = 0.05) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Left Ventricular Diameter (mm)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  stat_summary(fun.y = mean, geom = "point", shape = 5, size  = 3, color = "#9A2617") + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue") +
  jn_theme() +
  theme(axis.title.x = element_blank())
LV_Diam_systole_15


LVAW_dia_15 <- ggplot(data = EchoC_data[EchoC_data$Group == "September",], aes(x = Genotype, y = LVAW_d)) + 
  geom_violin(scale = "area", adjust = 1, aes(color = Genotype), show.legend = FALSE) + 
  scale_color_manual(values = c("Wild_Type"="Black", "Mutant"="Red")) +
  geom_point(size = 0.5, height = 0, width = 0.05) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Left Ventricular Anterior Wall Diameter (mm)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  stat_summary(fun.y = mean, geom = "point", shape = 5, size  = 3, color = "#9A2617") + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue") +
  jn_theme() +
  theme(axis.title.x = element_blank())
LVAW_dia_15

Ejection_fraction_15 <- ggplot(data = EchoC_data[EchoC_data$Group == "September",], aes(x = Genotype, y = EF)) + 
  geom_violin(scale = "area", adjust = 1, aes(color = Genotype), show.legend = FALSE) + 
  scale_color_manual(values = c("Wild_Type"="Black", "Mutant"="Red")) +
  geom_point(size = 0.5, height = 0, width = 0.05) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Ejection Fraction (%)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  stat_summary(fun.y = mean, geom = "point", shape = 5, size  = 3, color = "#9A2617") + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue") +
  jn_theme() +
  theme(axis.title.x = element_blank())
Ejection_fraction_15


LVOT_Mean_grad_15 <- ggplot(data = EchoC_data[EchoC_data$Group == "September",], aes(x = Genotype, y = LVOT_mean_grad)) + 
  geom_violin(scale = "area", adjust = 1, aes(color = Genotype), show.legend = FALSE) + 
  scale_color_manual(values = c("Wild_Type"="Black", "Mutant"="Red")) +
  geom_point(size = 0.5, height = 0, width = 0.05) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Mean Gradient (mmHg)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  stat_summary(fun.y = mean, geom = "point", shape = 5, size  = 3, color = "#9A2617") + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue") +
  jn_theme() +
  theme(axis.title.x = element_blank())
LVOT_Mean_grad_15

Aorta_sys_15<- ggplot(data = EchoC_data[EchoC_data$Group == "September",], aes(x = Genotype, y = Aor_sys)) + 
  geom_violin(scale = "area", adjust = 1, aes(color = Genotype), show.legend = FALSE) + 
  scale_color_manual(values = c("Wild_Type"="Black", "Mutant"="Red")) +
  geom_point(size = 0.5, height = 0, width = 0.05) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Aorta Diameter (mm)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  stat_summary(fun.y = mean, geom = "point", shape = 5, size  = 3, color = "#9A2617") + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue") +
  jn_theme() +
  theme(axis.title.x = element_blank())
Aorta_sys_15
```


Row {.tabset data-height=400}
------------------------------------------------------------------------------

### Tidied paired violin plots 
```{r plots}

jn_theme <- function(){
  theme_bw() +
    theme(axis.text = element_text(size = 14, color = "Black"),
          axis.title = element_text(size = 16),
          panel.grid.minor = element_blank(),
          strip.text = element_text(size=14),
          strip.background = element_blank(),
          plot.title = element_text(size = 20, hjust = 0.5),
          panel.grid.major.x = element_blank())
}


LV_EF <- ggplot(EchoC_data, aes(x = Genotype, y = EF, fill = Group)) + 
  geom_violin(scale = "area", adjust = 1) + 
  scale_fill_manual(values=c("#999999", "#D4AF37", "#56B4E9")) +
  theme_bw() + 
  geom_point(position=position_dodge(width = 0.9), aes(group = Group), size = .5, height = 0, width = 0.05, show.legend = FALSE) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Ejection Fraction %") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  ggtitle("Ejection Fraction Variation") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_summary(fun.y = mean, geom="point", shape = 18, size  = 3, color = "red", position = position_dodge(width = 0.9), show.legend = FALSE) + 
  stat_summary(fun.y=median,  geom = "point", shape = 3, size = 3, color = "blue", position = position_dodge(width = 0.9), show.legend = FALSE) +
  jn_theme()
LV_EF


LVOT_Mean_grad <- ggplot(EchoC_data, aes(x = Genotype, y = LVOT_mean_grad, fill = Group)) + 
  geom_violin(scale = "area", adjust = 1) + 
  scale_fill_manual(values=c("#999999", "#D4AF37", "#56B4E9")) +
  theme_bw() + 
  geom_point(position=position_dodge(width = 0.9), aes(group = Group), size = .5, height = 0, width = 0.05, show.legend = FALSE) + 
  ylab("Left Ventricular Outflow Tract Mean Gradient") + #UNITS???
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  ggtitle("Left Ventricular Outflow Tract Mean Gradient Variation") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_summary(fun.y = mean, geom="point", shape = 18, size  = 3, color = "red", position = position_dodge(width = 0.9), show.legend = FALSE) + 
  stat_summary(fun.y=median,  geom = "point", shape = 3, size = 3, color = "blue", position = position_dodge(width = 0.9), show.legend = FALSE) +
  jn_theme()
LVOT_Mean_grad


Aorta_sys <- ggplot(EchoC_data, aes(x = Genotype, y = Aor_sys, fill = Group)) + 
  geom_violin(scale = "area", adjust = 1) + 
  scale_fill_manual(values=c("#999999", "#D4AF37", "#56B4E9")) +
  theme_bw() + 
  geom_point(position=position_dodge(width = 0.9), aes(group = Group), size = .5, height = 0, width = 0.05, show.legend = FALSE) + 
  ylab("Aorta Diameter (mm)") + #UNITS???
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  ggtitle("Variations in Aortic Diameter in Systole") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_summary(fun.y = mean, geom="point", shape = 18, size  = 3, color = "red", position = position_dodge(width = 0.9), show.legend = FALSE) + 
  stat_summary(fun.y=median,  geom = "point", shape = 3, size = 3, color = "blue", position = position_dodge(width = 0.9), show.legend = FALSE) +
  jn_theme()
Aorta_sys

```





Interactive Plots {data-orientation=columns data-icon="fa-eye"}
===================================== 

Column {data-width=500}
------------------------------------------------------------------------------

### Using plotly

```{r interactive}

jn_theme <- function(){
  theme_bw() +
    theme(axis.text = element_text(size = 14, color = "Black"),
          axis.title = element_text(size = 16),
          panel.grid.minor = element_blank(),
          strip.text = element_text(size=14),
          strip.background = element_blank(),
          plot.title = element_text(size = 20, hjust = 0.5),
          panel.grid.major.x = element_blank())
}


LV_Sys_Vol <- ggplot(EchoC_data, aes(x = Genotype, y = LV_Vol_s, fill = Group, label = ID)) + 
  geom_violin() + 
  xlim(0, 20) +
  scale_fill_manual(values=c("#999999", "#D4AF37", "#56B4E9")) +
  geom_point(position=position_dodge(width = .9), aes(group = Group), size = .5, height = 0, width = 0.05, show.legend = FALSE) + #can use geom_jitter to introduce a small amount of random variation to the location of each point to handle overplott
  ylab("Left Ventricular Systolic Volume (μL)") + 
  scale_x_discrete(limits = c("Wild_Type", "Mutant")) + 
  ggtitle("Systolic Volume Variation") + 
  stat_summary(fun.y = mean, geom = "point", shape = 18, size  = 3, color = "red", position = position_dodge(width = 0.9), show.legend = FALSE) + 
  stat_summary(fun.y = median,  geom = "point", shape = 3, size = 3, color = "blue", position = position_dodge(width = 0.9), show.legend = FALSE) +
  jn_theme()

LV_Sys_Vol




#Messing around:
library(plotly)
ggplotly(LV_Sys_Vol, tooltip = c("ID", "LV_Vol_s"))%>%layout(violinmode = 'group', violingap = 1, violingroupgap = 1)

```


About {data-orientation=rows data-icon="fa-comment-alt"}
===================================== 

Row {data-height=1000}
------------------------------------------------------------------------------

Dashboard made by Jacob Noeker:

* Made for Data Visualization in R
* Thank you to Abhijit Dasgupta

## Abhijit's Comments

A nice presentation, using flexdashboard to demonstrate different types of 
plots. My main criticism would be on the first page, where you're presenting 
too much raw information, which makes it harder to read and comprehend quickly. 
Raw results from tests should __never__ be presented, and `broom::tidy` is your friend
for that. You might also want to think about the size of the axis titles on page 2, since
some of them are running into each other. 

One technical comment is that Genotype should be a factor, otherwise I can't
see it's frequencies in the summary panel. On the flip side, ID shouldn't really
be reported. 
