---
title: "Final Project"
author: Rachael Philips
date: May 21, 2019
output: html_document
---

```{r setup, echo = T, message = F, warning = F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(ggpubr)
library(gridExtra)
library(cowplot)
```

## Background

Patients with a STAT1 gain-of-function (GOF) mutation are susceptible to chronic fungal infections.  STAT1 is a transcription factor downstream of multiple cytokine signaling pathways and regulates gene expression in response to various cytokines.  To determine how the STAT1-GOF mutation alters gene expression, isolated CD4 T cells were cultured under Th0 conditions for 3 days, with or without IL-27 and IL-6.

```{r Gene_expression}

GSE65621_table2_human_RPKM <- read.csv("GSE65621_table2_human_RPKM.csv")

glimpse(GSE65621_table2_human_RPKM)
```

## Data manipulation

Each sample (column) was mutated to add 1 to each variable and determine the fold-change between no cytokine (nc) and cytokine (IL-6 or IL-27).  Then each gene was averaged according to sample type (healthy control (HC) or STAT1-GOF mutation (S1GOF)).

```{r manipulation, echo=TRUE}

All_RPKM_plus1 <- GSE65621_table2_human_RPKM %>% 
  mutate(mRNA_HC1_nc_3d_693 = mRNA_HC1_nc_3d_693 + 1) %>% 
  mutate(mRNA_HC2_nc_3d_786 = mRNA_HC2_nc_3d_786 + 1) %>% 
  mutate(mRNA_HC3_nc_3d_789 = mRNA_HC3_nc_3d_789 + 1) %>% 
  mutate(mRNA_HC4_nc_3d_835 = mRNA_HC4_nc_3d_835 + 1) %>% 
  mutate(mRNA_HC1_IL6_3d_694 = mRNA_HC1_IL6_3d_694 + 1) %>% 
  mutate(mRNA_HC2_IL6_3d_787 = mRNA_HC2_IL6_3d_787 + 1) %>% 
  mutate(mRNA_HC3_IL6_3d_790 = mRNA_HC3_IL6_3d_790 + 1) %>% 
  mutate(mRNA_HC4_IL6_3d_836 = mRNA_HC4_IL6_3d_836 + 1) %>% 
  mutate(mRNA_HC1_IL27_3d_695 = mRNA_HC1_IL27_3d_695 + 1) %>% 
  mutate(mRNA_HC2_IL27_3d_788 = mRNA_HC2_IL27_3d_788 + 1) %>% 
  mutate(mRNA_HC3_IL27_3d_791 = mRNA_HC3_IL27_3d_791 + 1) %>% 
  mutate(mRNA_HC4_IL27_3d_837 = mRNA_HC4_IL27_3d_837 + 1) %>% 
  mutate(mRNA_S1GOF1_nc_3d_621 = mRNA_S1GOF1_nc_3d_621 + 1) %>% 
  mutate(mRNA_S1GOF2_nc_3d_795 = mRNA_S1GOF2_nc_3d_795 + 1) %>% 
  mutate(mRNA_S1GOF3_nc_3d_838 = mRNA_S1GOF3_nc_3d_838 + 1) %>% 
  mutate(mRNA_S1GOF4_nc_3d_841 = mRNA_S1GOF4_nc_3d_841 + 1) %>%
  mutate(mRNA_S1GOF5_nc_3d_844 = mRNA_S1GOF5_nc_3d_844 + 1) %>%
  mutate(mRNA_S1GOF1_IL6_3d_622 = mRNA_S1GOF1_IL6_3d_622 + 1) %>% 
  mutate(mRNA_S1GOF2_IL6_3d_796 = mRNA_S1GOF2_IL6_3d_796 + 1) %>% 
  mutate(mRNA_S1GOF3_IL6_3d_839 = mRNA_S1GOF3_IL6_3d_839 + 1) %>% 
  mutate(mRNA_S1GOF4_IL6_3d_842 = mRNA_S1GOF4_IL6_3d_842 + 1) %>%
  mutate(mRNA_S1GOF5_IL6_3d_845 = mRNA_S1GOF5_IL6_3d_845 + 1) %>%
  mutate(mRNA_S1GOF1_IL27_3d_623 = mRNA_S1GOF1_IL27_3d_623 + 1) %>% 
  mutate(mRNA_S1GOF2_IL27_3d_797 = mRNA_S1GOF2_IL27_3d_797 + 1) %>% 
  mutate(mRNA_S1GOF3_IL27_3d_840 = mRNA_S1GOF3_IL27_3d_840 + 1) %>% 
  mutate(mRNA_S1GOF4_IL27_3d_843 = mRNA_S1GOF4_IL27_3d_843 + 1) %>%
  mutate(mRNA_S1GOF5_IL27_3d_846 = mRNA_S1GOF5_IL27_3d_846 + 1)

#AD This could have been replaced by 
#AD All_RPKM_plus1_AD <- GSE65621_table2_human_RPKM %>% 
#AD   mutate_at(vars(starts_with("mRNA")), function(x) {x + 1})
#AD 
#AD To check
#AD all(All_RPKM_plus1 == All_RPKM_plus1_AD)

IL6vNC <- All_RPKM_plus1 %>% 
  mutate(HC1_IL6vNC = mRNA_HC1_IL6_3d_694 / mRNA_HC1_nc_3d_693,
         HC2_IL6vNC = mRNA_HC2_IL6_3d_787 / mRNA_HC2_nc_3d_786,
         HC3_IL6vNC = mRNA_HC3_IL6_3d_790 / mRNA_HC3_nc_3d_789,
         HC4_IL6vNC = mRNA_HC4_IL6_3d_836 / mRNA_HC4_nc_3d_835,
         S1gof1_IL6vNC = mRNA_S1GOF1_IL6_3d_622 / mRNA_S1GOF1_nc_3d_621,
         S1gof2_IL6vNC = mRNA_S1GOF2_IL6_3d_796 / mRNA_S1GOF2_nc_3d_795,
         S1gof3_IL6vNC = mRNA_S1GOF3_IL6_3d_839 / mRNA_S1GOF3_nc_3d_838,
         S1gof4_IL6vNC = mRNA_S1GOF4_IL6_3d_842 / mRNA_S1GOF4_nc_3d_841,
         S1gof5_IL6vNC = mRNA_S1GOF5_IL6_3d_845 / mRNA_S1GOF5_nc_3d_844) %>% 
  select(-1, -3:-29)

#AD IL5vNC_AD <- All_RPKM_plus1 %>% 
#AD   mutate(HC1_IL6vNC = mRNA_HC1_IL6_3d_694 / mRNA_HC1_nc_3d_693,
#AD          HC2_IL6vNC = mRNA_HC2_IL6_3d_787 / mRNA_HC2_nc_3d_786,
#AD          HC3_IL6vNC = mRNA_HC3_IL6_3d_790 / mRNA_HC3_nc_3d_789,
#AD          HC4_IL6vNC = mRNA_HC4_IL6_3d_836 / mRNA_HC4_nc_3d_835,
#AD          S1gof1_IL6vNC = mRNA_S1GOF1_IL6_3d_622 / mRNA_S1GOF1_nc_3d_621,
#AD          S1gof2_IL6vNC = mRNA_S1GOF2_IL6_3d_796 / mRNA_S1GOF2_nc_3d_795,
#AD          S1gof3_IL6vNC = mRNA_S1GOF3_IL6_3d_839 / mRNA_S1GOF3_nc_3d_838,
#AD          S1gof4_IL6vNC = mRNA_S1GOF4_IL6_3d_842 / mRNA_S1GOF4_nc_3d_841,
#AD          S1gof5_IL6vNC = mRNA_S1GOF5_IL6_3d_845 / mRNA_S1GOF5_nc_3d_844) %>% 
#AD   select(-starts_with("mRNA"), -RefSeq_ID)

IL27vNC <- All_RPKM_plus1 %>% 
  mutate(HC1_IL27vNC = mRNA_HC1_IL27_3d_695 / mRNA_HC1_nc_3d_693,
         HC2_IL27vNC = mRNA_HC2_IL27_3d_788 / mRNA_HC2_nc_3d_786,
         HC3_IL27vNC = mRNA_HC3_IL27_3d_791 / mRNA_HC3_nc_3d_789,
         HC4_IL27vNC = mRNA_HC4_IL27_3d_837 / mRNA_HC4_nc_3d_835,
         S1gof1_IL27vNC = mRNA_S1GOF1_IL27_3d_623 / mRNA_S1GOF1_nc_3d_621,
         S1gof2_IL27vNC = mRNA_S1GOF2_IL27_3d_797 / mRNA_S1GOF2_nc_3d_795,
         S1gof3_IL27vNC = mRNA_S1GOF3_IL27_3d_840 / mRNA_S1GOF3_nc_3d_838,
         S1gof4_IL27vNC = mRNA_S1GOF4_IL27_3d_843 / mRNA_S1GOF4_nc_3d_841,
         S1gof5_IL27vNC = mRNA_S1GOF5_IL27_3d_846 / mRNA_S1GOF5_nc_3d_844) %>% 
  select(-1, -3:-29)

#AD IL27vNC_AD <- All_RPKM_plus1 %>% 
#AD   mutate(HC1_IL27vNC = mRNA_HC1_IL27_3d_695 / mRNA_HC1_nc_3d_693,
#AD          HC2_IL27vNC = mRNA_HC2_IL27_3d_788 / mRNA_HC2_nc_3d_786,
#AD          HC3_IL27vNC = mRNA_HC3_IL27_3d_791 / mRNA_HC3_nc_3d_789,
#AD          HC4_IL27vNC = mRNA_HC4_IL27_3d_837 / mRNA_HC4_nc_3d_835,
#AD          S1gof1_IL27vNC = mRNA_S1GOF1_IL27_3d_623 / mRNA_S1GOF1_nc_3d_621,
#AD          S1gof2_IL27vNC = mRNA_S1GOF2_IL27_3d_797 / mRNA_S1GOF2_nc_3d_795,
#AD          S1gof3_IL27vNC = mRNA_S1GOF3_IL27_3d_840 / mRNA_S1GOF3_nc_3d_838,
#AD          S1gof4_IL27vNC = mRNA_S1GOF4_IL27_3d_843 / mRNA_S1GOF4_nc_3d_841,
#AD          S1gof5_IL27vNC = mRNA_S1GOF5_IL27_3d_846 / mRNA_S1GOF5_nc_3d_844) %>% 
#AD   select(-starts_with("mRNA"), -RefSeq_ID)


DF_IL6vNC <- IL6vNC %>% 
  mutate(HC_IL6vNC_means = rowMeans(.[,2:5])) %>% 
  mutate(S1gof_IL6vNC_means = rowMeans(.[,6:10]))

#AD DF_IL6vNC_AD <- IL6vNC %>% 
#AD   mutate(HC_IL6vNC_means = rowMeans(select(., starts_with("HC")))) %>% 
#AD   mutate(S1gof_IL6vNC_means = rowMeans(select(., starts_with("S1gof"))))
#AD all(DF_IL6vNC_AD == DF_IL6vNC)

DF_IL6vNC <- DF_IL6vNC %>%
  mutate(S1gofFC_larger = S1gof_IL6vNC_means / HC_IL6vNC_means) 

DF_IL6vNC <- DF_IL6vNC %>%
  mutate(group = case_when(  #AD Nice!!!
    S1gofFC_larger >= 1.5 ~ 1,
    S1gofFC_larger < 1.5 & S1gofFC_larger > 0.667 ~ 2,
    S1gofFC_larger <= 0.667 ~ 3))

DF_IL27vNC <- IL27vNC %>% 
  mutate(HC_IL27vNC_means = rowMeans(.[,2:5])) %>% 
  mutate(S1gof_IL27vNC_means = rowMeans(.[,6:10])) 

#AD DF_IL27vNC_AD <- IL27vNC %>% 
#AD   mutate(HC_IL27vNC_means = rowMeans(select(., starts_with('HC')))) %>% 
#AD   mutate(S1gof_IL27vNC_means = rowMeans(select(., starts_with('S1gof'))))
#AD all(DF_IL27vNC_AD == DF_IL27vNC)

```

## Plots

Each plot shows the change in gene expression after stilumation with IL-27 or IL-6 from Healthy Controls and STAT1-GOF patients.  

```{r plots, echo=TRUE}

pointsToLabel <- c("GZMB", "ADM", "GNA15", "NKG7", "CCL4", "CXCL10", "TBX21",
                   "DDIT4", "NRN1", "CSF2", "CCR4", "NPW")

# plots for IL-27

plot_IL27 <- ggplot(data = DF_IL27vNC, aes(x = log2(HC_IL27vNC_means), y = log2(S1gof_IL27vNC_means))) +
  geom_point(color = "cornflowerblue") +
  geom_smooth() +
  theme_bw() +
  xlab("Healthy Control (log2FC)") +
  ylab("STAT1-GOF Patient (log2FC)") +
  ggtitle("IL-27 vs No Cytokine") +
  theme(plot.title = element_text(color = "#3366FF", vjust = 0.2)) +
  geom_text_repel(aes(label = Gene_symbol),
                  size = 4,
                  force = 50,
                  box.padding = unit(0.5, "lines"),
                  point.padding = unit(0.5, "lines"),
                  arrow = arrow(length = unit(0.015, "npc")),
                  color = "gray20",
                  data = DF_IL27vNC %>%  #AD Nice!!
                    filter(Gene_symbol %in% pointsToLabel)) +
  xlim(-3, 4) + ylim(-5, 6)

# plots for IL-6

plot_IL6 <- ggplot(data = DF_IL6vNC, aes(x = log2(HC_IL6vNC_means), y = log2(S1gof_IL6vNC_means))) +
  geom_point(color = "coral") +
  geom_smooth(color = "coral4") +
  theme_bw() +
  xlab("Healthy Control (log2FC)") +
  ylab("STAT1-GOF Patient (log2FC)") +
  ggtitle("IL-6 vs No Cytokine") +
  theme(plot.title = element_text(color = "coral4", vjust = 0.2)) +
  geom_text_repel(aes(label = Gene_symbol),
                  size = 4,
                  force = 50,
                  box.padding = unit(0.5, "lines"),
                  point.padding = unit(0.5, "lines"),
                  arrow = arrow(length = unit(0.015, "npc")),
                  color = "gray20",
                  data = DF_IL6vNC %>% 
                    filter(Gene_symbol %in% pointsToLabel)) +
  xlim(-3, 4) + ylim(-5, 6)

# plot multiple panels

MyFigure <- ggarrange (plot_IL27, plot_IL6,  #AD Nice!!
                       ncol = 2, nrow = 1)

annotate_figure(MyFigure, 
                top = text_grob("Effect of STAT1 gain-of-function (GOF) mutation on gene expression",
                                size = 14, face = "bold"))

```


## Abhijit's comments

Generally a beautiful job. You've actually taught me some of the options in 
`geom_text_repel`. One thing I prefer generally is that selection happens by 
name rather than by position/index. That way the selection is clear when I 
look at the code again. Which also means that I'm careful naming my variables 
(as you've been) so that they are easily searchable using `starts_with`, `ends_with`
or `contains`. I've also cleaned up your presentation Rmd so that we don't see 
quite as much messages that are irrelevant to the presentation. 

Since your data is in the same folder as your code, don't put the full 
pathname in, since, for example, I couldn't directly read the file. The better
method on your computer or even here is to use the `here` package for file names. 
See [this presentation](https://speakerdeck.com/jennybc/zen-and-the-art-of-workflow-maintenance?slide=47) or really anything by Jenny Bryan.
